---
title: "Practical Machine Learning Project"
author: "Wouter van Amsterdam"
date: "1 September 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

The problem defined for this project is how to classify proper weight lifting with on-body sensors
See also: http://groupware.les.inf.puc-rio.br/har

This markdown document will guide you through the process of model building and validation, and show the acheived results

NB I will use the data.table package for my data handling, which uses a syntax for subsetting that's a bit different from data.frames
See: https://cran.r-project.org/web/packages/data.table/vignettes/datatable-intro.pdf

## Downloading and importing data

## Download data

```{r download_data, echo = T, results='hide', cache=T}
url <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
download.file(url, destfile = 'data/pml-training.csv')
url2 <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
download.file(url2, destfile = 'data/pml-testing.csv')
```

## Import data into R

```{r import_data, echo = T, results = 'hide', cache = T}
library(data.table)
training <- data.table(read.csv("data/pml-training.csv", header = T))
testing  <- data.table(read.csv("data/pml-testing.csv", header = T))
```

## Data exploration and curation

```{r, echo = T, results='hide'}
library(data.table)
head(training)
head(testing)
colSums(is.na(training))
```
```{r, echo = T}
colSums(is.na(testing))/nrow(testing)
training[, lapply(.SD, table), .SDcols = c("user_name", "new_window", "num_window")]
training[, table(user_name, new_window)]
training[, table(num_window)]
```

Take a better look at the timestamps. CVTD_timestamp is the date and time, 
raw_timestamp_part_2 probably contains the millisecond information, and 
raw_timestamp_part_1 the second information (last two digits)
```{r timestamp, echo = T}
training[, head(cbind(raw_timestamp_part_1, raw_timestamp_part_2, 
                      as.character(cvtd_timestamp)), 30)]

```



Looks like some variables are 100% missing in the testing data. They won't provide any information in model testing, let's throw 'm out

```{r, echo = T, results = 'hide'}
empty_vars <- names(testing)[colSums(is.na(testing)) == nrow(testing)]
training <- training[, .SD, .SDcols = !names(training) %in% empty_vars]
testing  <- testing[, .SD, .SDcols = !names(testing) %in% empty_vars]
```


Now that we've kicked out the empty variables, let's look at the 60 remaining variables. 'X' in this dataset is just a rownumber, and can be excluded

```{r, echo = T, results='hide'}
training <- training[, .SD, .SDcols = -c("X")]
testing <- testing[, .SD, .SDcols = -c("X")]
colSums(is.na(training))

library(xtable)
## reminder to use xtable for html pretty tables
```

That's it!.

